From heartbeat to data -- Using wearable fitness trackers as an affordable approach to assess teacher stress

### 2024-11-07 02_paper/02_study/01_fitbit_paper/Submission/FitBit_submission_template.tex
literaturverzeichnis

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage[]{natbib}
\bibliographystyle{plainnat}

@@

\setcitestyle{sort&compress}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{caption}

@@

The teaching profession is one of the most stressful professions, with
teachers facing a host of stressors during their everyday work
\citep{smith2000, herman2020, schult2014belastet}. To better understand
mechanisms in teacher stress, there is a growing research interest in
physiological measures such as heart rate (HR) as online measures of
teachers' stress during teaching
\citep{karner2021teachers, wettstein2020ambulatory}. For example, it has
been shown that teacher-centered activities and typical
classroom-related stressors increase teacher HR during teaching
activities
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021, huang2022class}.
However, previous studies have often relied on expensive and obtrusive
electrocardiographs (ECG). Modern fitness trackers represent a
non-invasive and convenient alternative \citep{ferguson2015}.

Classroom disruptions are a major stressor in teachers' daily work
\citep{boyle1995structural, aloe2014multivariate}, and learning how to
deal with them is an important aspect of professional expertise
\citep{wolff2015keeping}. According to \citet{lazarus1990theory}
transactional model of stress and coping, the experience of stress in
response to stressors such as classroom disruptions depends on the
teacher´s subjective appraisal, which, in turn, depends on their coping
resources, such as their professional knowledge. The resulting stress
response has a psychological, physiological, or behavioral dimension
\citep{kyriacou1978}. Therefore, in order to better understand how
classroom stressors affect teachers' stress response, subjective
self-reports should be accompanied by objective, physiological measures
\citep{wettstein2021}. Teachers' use of wrist-worn fitness trackers in
educational research provides fine-grained, in vivo data, allowing
researchers as well as teachers themselves to monitor their
physiological stress response continuously during teaching, across
settings, and at low costs. Being able to monitor, and eventually
counteract, teacher stress levels appears particularly relevant given
the profession's generally high stress levels and associated negative
health effects \citep{johnson2005experience, montgomery2005meta}. To
harness this potential, the present study explored the use of
wrist-based fitness trackers as a tool to assess teachers' HR, as an
indicator of stress, before, during, and after a teaching session during
which typical, potentially stressful, classroom disruptions occurred.
Further, we explored to what extent teachers' subjective appraisals of
classroom disruptions and their teaching experience predicted teacher
stress as assessed by the fitness tracker.

@@

Fitness trackers provide data on physical activity and cardiovascular
parameters such as HR, supporting personalized fitness goals
\citep{nuss2021effects} and stress management \citep{hao2018chrv}. They
can be used as ubiquitous, low-cost, and unintrusive data collection
instruments \citep{godfrey2018z}, and their wide-spread use and everyday
availability align with the increasing popularity and acceptance of
wearables among the general population \citep{peng2022acceptance}. In
contrast to self-reported questionnaires on stress
\citep{chaplain2008, liu2020} that are prone to biases like social
desirability \citep{razavi2001self} or recall errors
\citep{van2016accuracy}, fitness trackers, as ambulatory assessment
methods \citep{trull2013ambulatory, wettstein2020ambulatory}, offer more
objective insights into teachers' stress levels by monitoring teachers'
physiological stress responses without disrupting teaching
\citep{donker2018, runge2020}.

One important health parameter assessed by nearly all wrist-worn fitness
trackers is heart rate \citep{scalise2018wearables}. HR indicates the
number of heartbeats within one minute and is typically expressed as
beats per minute (BPM)
\citep{berntson2007cardiovascular, hottenrott2007}. At rest, the average
HR of adults typically ranges between 60 and 80 BPM
\citep{sammito2015guideline}. HR can be detected and measured in
different ways using sensors, such as electrocardiography (ECG) or
photoplethysmography (PPG) \citep{mukhopadhyay2017wearable}. While ECG
sensors offer precise measurements by detecting the heart's electrical
activity, their intrusive nature and requirement of direct skin contact
may limit their suitability \citep{kranjec2014non}, particularly in
educational settings. In contrast, photoplethysmography (PPG) is a
rather uncomplicated and inexpensive technique to measure HR, commonly
found in commercially available fitness trackers
\citep{castaneda2018review}. This optical method assesses HR by flashing
green or red lights to measure changes in blood volume in the
capillaries of the skin \citep{allen2007photoplethysmography}.

Physiologically, HR is regulated by the sympathetic and parasympathetic
nervous systems \citep{pham2021}. An increase in sympathetic activity
results in HR being sped up (``fight or flight'' response)
\citep{taelman2009influence}; whereas an increase in parasympathetic
activity results in HR being slowed down (``rest and digest'' response,
\citet{battipaglia2015}). Stress or mental and physical strain directly
increases HR \citep{custodis2014heart, sachs2014}. Therefore, an
increase in HR can be regarded as an indicator of increasing stress, and
a decrease as an indicator of relaxation and ease \citep{kyriacou1978}.
Thus, fitness trackers offer low-cost and unobtrusive access to
psychological stress data.

@@

Prior research, typically using ECG methods, has shown that changes in
teachers' HR can be mapped onto stressors they experience during
teaching. For example, teachers' HR tends to increase when teachers take
an exposed position in student-teacher interaction
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
\citet{sperka1995} for example recorded the HR of 16 pre-service
teachers during their first lesson and showed that teachers' HR
increased significantly during teaching. The activation was particularly
prominent at the beginning of the lesson and decreased over the course
of the lesson. The authors suggested that pre-service teachers'
proactive coping strategies, such as actively managing student
interactions, helped lower their HR levels. Other ECG studies identified
typical stressors predicting increases in HR, such as class size
\citep{huang2022class}, or low student engagement and motivation
\citep{junker2021}. \citet{junker2021} recorded the HR of 40 teachers
during a real classroom lesson. Again, teacher stress, induced by
factors such as low student engagement (e.g., lack of motivation or
interest in tasks) or teacher-centered activities (e.g., teacher-focused
classroom activities), resulted in elevated HR.

More recent studies have begun using wrist-worn fitness trackers to
investigate HR trends in instructional settings
\citep{Darnell2019, chalmers2021}. \citet{Darnell2019} measured the HR
of 15 medical college students listening to lectures, using wrist-worn
devices. The analysis revealed a constant decrease in HR throughout the
lecture, with HR peaks during more interactive learning phases.
\citet{chalmers2021} used HR data from a fitness tracker to identify
physiological changes during stress-inducing tasks (i.e., the Trier
Social Stress Test; \citet{kirschbaum1993trier}). Average HR increased
significantly from the resting to the stress inducing phases. Even
though the participants of these previous studies
\citep{Darnell2019, chalmers2021} were not teachers but learners, the
results are relevant for studying teacher stress as they demonstrated
that HR can be effectively recorded using fitness trackers over the
course of a whole learning unit, and HR changes are in line with the
occurrence of activating or stress-inducing tasks.

To the best of our knowledge, only one study has directly assessed
teachers' HR during teaching using a wrist-worn fitness tracker:
\citet{runge2020} assessed HR as an indicator of stress in four
in-service teachers during authentic lessons. They used fitness
trackers' recordings to create a profile for each teacher, with the aim
of differentiating between teachers reporting higher vs.~lower levels of
stress. It was found that the combination of a high HR, a high number of
steps, and short sleep duration was characteristic for teachers
reporting high stress levels. However, it should be noted that the
generalizability of these results is limited due to the small sample
size.

In summary, previous studies have revealed that teachers' (and
students') HR changes depend on their activities and the stressors they
experience, with an increase in HR already before the expected stressors
occur, and with peaks in activating phases
\citep{Darnell2019, chalmers2021}. For teachers, teacher-centered phases
led to an increase in HR
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021}.
However, there is a lack of studies using teacher-worn fitness trackers
in larger samples, exploring the feasibility of this convenient tool for
researching links between teachers' HR and stress inducing events and
mechanisms.

@@

According to \citet{kyriacou1978}, teacher stress can be defined as a
negative affective response, typically accompanied by physiological
changes such as increased HR, triggered by job-related demands, and
perceived as threatening to one's self-esteem or well-being. Coping
mechanisms help to reduce the perceived threat. Kyriacou's definition of
teacher stress (see \citet{kyriacou1978} and, for a more recent
adaptation, \citet{van2006stress}) is based on the transactional stress
model by Lazarus and colleagues
\citep{lazarus1966psychological, lazarus1990theory}, which highlights
the interaction between an individual and the environment, whereby
stress refers to a person's subjective reaction to an event (a stressor)
that exceeds the person's adaptive resources.

Fig.\ref{fig.1} shows, in a simplified way, how classroom events affect
teachers' stress level, according to the adaptation of the Lazarus model
to teacher stress proposed by \citet{van2006stress}: When potential
stressors (e.g., classroom disruptions) occur during teaching (1),
teachers intuitively judge how disruptive the event is in a primary
appraisal (2). If potential stressors are judged as threatening, i.e.,
as actual stressors (3), teachers consider whether they have sufficient
resources for coping with the stressors (4). Teachers utilize these
resources in trying to cope with the stressors, e.g., by employing
classroom management strategies (5). In cases when coping fails, stress
ensues, often accompanied by physiological reactions like increased HR
(6). As part of the coping process, and dependent on its outcomes,
teachers re-evaluate the stressor (7).

As shown in Fig.\ref{fig.1}, teachers' primary and secondary appraisal
as well as coping attempts are influenced by professional experience. As
professional experience grows, teachers develop cognitive scripts for
managing classroom events, resulting in more complex and problem-focused
classroom management skills \citep{wolff2021classroom}, and thus more
effective coping and less stress. Empirically, classroom management
skills and problem-focused coping styles are linked to fewer instances
of emotional exhaustion \citep{maslach2001job, clunies2008self}. Novices
in the teaching profession, on the other hand, face considerable stress
and often feel overwhelmed by the demands of teaching
\citep{ophardt2017klassenmanagement, wolff2015keeping, klusmann2012berufliche}
with many leaving the profession within the first five years
\citep{ingersoll2003}. Accordingly, when resources are lacking and
coping fails, negative consequences for health (e.g., burnout) and for
work (e.g., high turnover rates) can arise
\citep{jalongo2006, unterbrink2007, aloe2014multivariate}, highlighting
the importance of professional expertise for managing teacher stress
\citep{fisher2011}.

@@

To measure teachers' HR, we used the wrist-based fitness tracker Fitbit®
Charge 4. In line with the manufacturer's instructions \citep{fitbitnd},
the device was attached to the participants' nondominant hand, a
finger's width above the wrist bone. The tracker works by flashing green
LEDs hundreds of times per second, using light-sensitive photodiodes to
catch the reflected light, to calculate the volume changes in the
capillaries. From this, the tracker calculated the heart beats per
minute. HR measurements were generated at least every 15
seconds\footnote{The fluctuations in the number of seconds in which the
  HR was measured are due to the participants' movements, meaning that
  the device could not measure the HR every second.}. The raw data
contained the estimated HR in BPM for each time stamp. To account for
individual differences in the baseline HR, we also calculated
z-standardized HR values based on individual means, i.e., at the subject
level of \(n\) = 81 participants (standardized HR).

Since we aimed to keep measurement intervals comparable between study
phases, we aggregated HR over a representative 10-minute interval within
each phase (cf.~Fig.\ref{fig.2}). Previous research has indicated that
10-minute intervals are a useful duration for analyzing PPG data
\citep{lu2008can}. The intervals were selected based on the following
rules: The \emph{pre-teaching interval} (\(I_1\)) comprised the first 10
minutes after the fitness tracker had been put on. The \emph{teaching
interval} (\(I_2\)) started two minutes after the lesson had started.
This interval was of the highest relevance to our study. We explicitly
chose an early 10-minute interval within the teaching phase, as previous
studies revealed that the beginning of a lesson is most demanding and
potentially stressful with regards to teacher-student interaction
\citep{donker2018, claessens2017positive}. The \emph{post-teaching
interval} (\(I_3\)) started immediately after the end of the teaching
unit. The \emph{interview interval} (\(I_4\)) was defined as the mid-10
minutes between the end of the teaching unit and the time point when the
fitness tracker was taken off. All participants were being interviewed
during this interval. The end interval (\(I_5\)) comprised the last 10
minutes before the fitness tracker was taken off.

@@

We conducted all analyses with R \citep{RStudio2020}. Graphics were
created using ggplot2 \citep{ggplot2}.

To enable visual inspection of HR trends, we displayed smoothed teacher
HR over the course of the recording.\footnote{The curve was smoothed
  using the geom\_smooth() function from the ggplot2 package in R
  \citep{ggplot2} based on the smoothing method LOESS (Locally Estimated
  Scatterplot Smoothing). This method fits a polynomial surface
  determined by one or more numerical predictors, using local fitting.}
We visually compared unstandardized and standardized HR trends over the
two-hour recording period.\footnote{Note that the study exceeded the
  planned duration of two hours for a few participants. To avoid
  distortions when mapping the HR over the course of the study (see
  Fig.\ref{fig.3}), the endpoint was set at two hours for all
  participants, even though data from later time points was used in the
  \emph{end interval} for a few participants.} For all further analyses,
we used standardized rather than unstandardized HR values.

We averaged each person's standardized HR over each of the five selected
intervals\footnote{We used the mean standardized HR instead of the mean
  intercept as we wanted to explain the mean HR of the entire intervals
  and not the HR at the very beginning of the interval (\(x\) = 0).},
resulting in one measure per person per interval. To test Hypothesis 1a,
we initially conducted a one-way ANOVA with repeated measures as an
omnibus test and then tested the mean differences between the
\emph{teaching interval} (\(I_2\)) and the other four intervals by
planned contrasts and inspection of effect size \(d\)
\citep{cohen1988statistical}.

For testing Hypothesis 1b, concerning HR changes within each interval,
we first conducted a linear estimation of the increase or decrease in
standardized HR values over time for each participant. To this end, we
used fixed intercept fixed slope regression models
\citep{gelman2006data} for each interval to estimate intercepts and
linear slopes for each individual, which were then averaged across
individuals.\footnote{Although this procedure does not account for
  nonmonotonic progressions in individual HR, a graphical evaluation
  revealed that the linear estimates corresponded well to the majority
  of the cases (see Fig.\ref{fig.a3} to \ref{fig.a7} in the
  supplementary material).} We tested Hypothesis 1b based on the
unstandardized estimates of mean slopes (one estimate per participant
per interval).

@@

These findings are in line with prior studies showing that teachers' HR
varies depending on their activities and encountered stressors with
increases during phases where teachers are in an exposed position
\citep{sperka1995, scheuch1997psychophysische, donker2018, junker2021},
as well as with findings showing how HR changes align with activating
events and stress-inducing tasks \citep{Darnell2019, chalmers2021}.

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
% TODO: Currently lineno needs to be loaded after amsmath because of conflict
% https://github.com/latex-lineno/lineno/issues/5
\usepackage{lineno} % add
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}

@@

% Pandoc citation processing
%From Pandoc 3.1.8
% definitions for citeproc citations
\NewDocumentCommand\citeproctext{}{}
\NewDocumentCommand\citeproc{mm}{%
  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
\makeatletter
 % allow citations to break across lines
 \let\@cite@ofmt\@firstofone
 % avoid brackets around text for \cite:
 \def\@biblabel#1{}
 \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
 {\begin{list}{}{%
  \setlength{\itemindent}{0pt}
  \setlength{\leftmargin}{0pt}
  \setlength{\parsep}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
   \setlength{\leftmargin}{\cslhangindent}
   \setlength{\itemindent}{-1\cslhangindent}
  \fi
  % set entry spacing
  \setlength{\itemsep}{#2\baselineskip}}}
 {\end{list}}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\bibliographystyle{elsarticle-num}
\setcitestyle{numbered}
\usepackage{tabularx}
\usepackage{threeparttable}
\usepackage{booktabs}
\usepackage{array}
\usepackage{caption}
\usepackage{tablefootnote}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{pdflscape}
\usepackage{longtable}
\usepackage{geometry}
\usepackage{caption}

@@

The teaching profession is one of the most stressful professions, with
teachers facing a host of stressors during their everyday work (Smith
2000; Herman et al. 2020; Schult, Münzer-Schrobildgen, and Sparfeldt
2014). To better understand mechanisms in teacher stress, there is a
growing research interest in physiological measures such as heart rate
(HR) as online measures of teachers' stress during teaching (Kärner and
Höning 2021; Wettstein et al. 2020). For example, it has been shown that
teacher-centered activities and typical classroom-related stressors
increase teacher HR during teaching activities (Sperka and Kittler 1995;
Scheuch and Knothe 1997; Donker, Van Gog, and Mainhard 2018; Junker,
Donker, and Mainhard 2021; Huang et al. 2022). However, previous studies
have often relied on expensive and obtrusive electrocardiographs (ECG).
Modern fitness trackers represent a non-invasive and convenient
alternative (Ferguson et al. 2015).

Classroom disruptions are a major stressor in teachers' daily work
(Boyle et al. 1995; Aloe et al. 2014), and learning how to deal with
them is an important aspect of professional expertise (Wolff et al.
2015). According to Lazarus (1990) transactional model of stress and
coping, the experience of stress in response to stressors such as
classroom disruptions depends on the teacher´s subjective appraisal,
which, in turn, depends on their coping resources, such as their
professional knowledge. The resulting stress response has a
psychological, physiological, or behavioral dimension (Kyriacou and
Sutcliffe 1978). Therefore, in order to better understand how classroom
stressors affect teachers' stress response, subjective self-reports
should be accompanied by objective, physiological measures (Wettstein et
al. 2021). Teachers' use of wrist-worn fitness trackers in educational
research provides fine-grained, in vivo data, allowing researchers as
well as teachers themselves to monitor their physiological stress
response continuously during teaching, across settings, and at low
costs. Being able to monitor, and eventually counteract, teacher stress
levels appears particularly relevant given the profession's generally
high stress levels and associated negative health effects (Johnson et
al. 2005; Montgomery and Rupp 2005). To harness this potential, the
present study explored the use of wrist-based fitness trackers as a tool
to assess teachers' HR, as an indicator of stress, before, during, and
after a teaching session during which typical, potentially stressful,
classroom disruptions occurred. Further, we explored to what extent
teachers' subjective appraisals of classroom disruptions and their
teaching experience predicted teacher stress as assessed by the fitness
tracker.

@@

Fitness trackers provide data on physical activity and cardiovascular
parameters such as HR, supporting personalized fitness goals (Nuss et
al. 2021) and stress management (Hao et al. 2017). They can be used as
ubiquitous, low-cost, and unintrusive data collection instruments
(Godfrey et al. 2018), and their wide-spread use and everyday
availability align with the increasing popularity and acceptance of
wearables among the general population (Peng et al. 2022). In contrast
to self-reported questionnaires on stress (Chaplain 2008; Liu and Yan
2020) that are prone to biases like social desirability (Razavi 2001) or
recall errors (Van den Bergh and Walentynowicz 2016), fitness trackers,
as ambulatory assessment methods (Trull and Ebner-Priemer 2013;
Wettstein et al. 2020), offer more objective insights into teachers'
stress levels by monitoring teachers' physiological stress responses
without disrupting teaching (Donker, Van Gog, and Mainhard 2018; Runge,
Haarman, and Fisher 2020).

One important health parameter assessed by nearly all wrist-worn fitness
trackers is heart rate (Scalise and Cosoli 2018). HR indicates the
number of heartbeats within one minute and is typically expressed as
beats per minute (BPM) (Berntson, Quigley, and Lozano 2007; Hottenrot
2007). At rest, the average HR of adults typically ranges between 60 and
80 BPM (Sammito et al. 2015). HR can be detected and measured in
different ways using sensors, such as electrocardiography (ECG) or
photoplethysmography (PPG) (Mukhopadhyay and Islam 2017). While ECG
sensors offer precise measurements by detecting the heart's electrical
activity, their intrusive nature and requirement of direct skin contact
may limit their suitability (Kranjec et al. 2014), particularly in
educational settings. In contrast, photoplethysmography (PPG) is a
rather uncomplicated and inexpensive technique to measure HR, commonly
found in commercially available fitness trackers (Castaneda et al.
2018). This optical method assesses HR by flashing green or red lights
to measure changes in blood volume in the capillaries of the skin (Allen
2007).

Physiologically, HR is regulated by the sympathetic and parasympathetic
nervous systems (Pham et al. 2021). An increase in sympathetic activity
results in HR being sped up (``fight or flight'' response) (Taelman et
al. 2009); whereas an increase in parasympathetic activity results in HR
being slowed down (``rest and digest'' response, Battipaglia and Lanza
(2015)). Stress or mental and physical strain directly increases HR
(Custodis et al. 2014; Sachs 2014). Therefore, an increase in HR can be
regarded as an indicator of increasing stress, and a decrease as an
indicator of relaxation and ease (Kyriacou and Sutcliffe 1978). Thus,
fitness trackers offer low-cost and unobtrusive access to psychological
stress data.

@@

Prior research, typically using ECG methods, has shown that changes in
teachers' HR can be mapped onto stressors they experience during
teaching. For example, teachers' HR tends to increase when teachers take
an exposed position in student-teacher interaction (Sperka and Kittler
1995; Scheuch and Knothe 1997; Donker, Van Gog, and Mainhard 2018;
Junker, Donker, and Mainhard 2021). Sperka and Kittler (1995) for
example recorded the HR of 16 pre-service teachers during their first
lesson and showed that teachers' HR increased significantly during
teaching. The activation was particularly prominent at the beginning of
the lesson and decreased over the course of the lesson. The authors
suggested that pre-service teachers' proactive coping strategies, such
as actively managing student interactions, helped lower their HR levels.
Other ECG studies identified typical stressors predicting increases in
HR, such as class size (Huang et al. 2022), or low student engagement
and motivation (Junker, Donker, and Mainhard 2021). Junker, Donker, and
Mainhard (2021) recorded the HR of 40 teachers during a real classroom
lesson. Again, teacher stress, induced by factors such as low student
engagement (e.g., lack of motivation or interest in tasks) or
teacher-centered activities (e.g., teacher-focused classroom
activities), resulted in elevated HR.

More recent studies have begun using wrist-worn fitness trackers to
investigate HR trends in instructional settings (Darnell and Krieg 2019;
Chalmers et al. 2021). Darnell and Krieg (2019) measured the HR of 15
medical college students listening to lectures, using wrist-worn
devices. The analysis revealed a constant decrease in HR throughout the
lecture, with HR peaks during more interactive learning phases. Chalmers
et al. (2021) used HR data from a fitness tracker to identify
physiological changes during stress-inducing tasks (i.e., the Trier
Social Stress Test; Kirschbaum, Pirke, and Hellhammer (1993)). Average
HR increased significantly from the resting to the stress inducing
phases. Even though the participants of these previous studies (Darnell
and Krieg 2019; Chalmers et al. 2021) were not teachers but learners,
the results are relevant for studying teacher stress as they
demonstrated that HR can be effectively recorded using fitness trackers
over the course of a whole learning unit, and HR changes are in line
with the occurrence of activating or stress-inducing tasks.

To the best of our knowledge, only one study has directly assessed
teachers' HR during teaching using a wrist-worn fitness tracker: Runge,
Haarman, and Fisher (2020) assessed HR as an indicator of stress in four
in-service teachers during authentic lessons. They used fitness
trackers' recordings to create a profile for each teacher, with the aim
of differentiating between teachers reporting higher vs.~lower levels of
stress. It was found that the combination of a high HR, a high number of
steps, and short sleep duration was characteristic for teachers
reporting high stress levels. However, it should be noted that the
generalizability of these results is limited due to the small sample
size.

In summary, previous studies have revealed that teachers' (and
students') HR changes depend on their activities and the stressors they
experience, with an increase in HR already before the expected stressors
occur, and with peaks in activating phases (Darnell and Krieg 2019;
Chalmers et al. 2021). For teachers, teacher-centered phases led to an
increase in HR (Sperka and Kittler 1995; Scheuch and Knothe 1997;
Donker, Van Gog, and Mainhard 2018; Junker, Donker, and Mainhard 2021).
However, there is a lack of studies using teacher-worn fitness trackers
in larger samples, exploring the feasibility of this convenient tool for
researching links between teachers' HR and stress inducing events and
mechanisms.

@@

According to Kyriacou and Sutcliffe (1978), teacher stress can be
defined as a negative affective response, typically accompanied by
physiological changes such as increased HR, triggered by job-related
demands, and perceived as threatening to one's self-esteem or
well-being. Coping mechanisms help to reduce the perceived threat.
Kyriacou's definition of teacher stress (see Kyriacou and Sutcliffe
(1978) and, for a more recent adaptation, Van Dick (2006)) is based on
the transactional stress model by Lazarus and colleagues (Lazarus 1966,
1990), which highlights the interaction between an individual and the
environment, whereby stress refers to a person's subjective reaction to
an event (a stressor) that exceeds the person's adaptive resources.

Fig.\ref{fig.1} shows, in a simplified way, how classroom events affect
teachers' stress level, according to the adaptation of the Lazarus model
to teacher stress proposed by Van Dick (2006): When potential stressors
(e.g., classroom disruptions) occur during teaching (1), teachers
intuitively judge how disruptive the event is in a primary appraisal
(2). If potential stressors are judged as threatening, i.e., as actual
stressors (3), teachers consider whether they have sufficient resources
for coping with the stressors (4). Teachers utilize these resources in
trying to cope with the stressors, e.g., by employing classroom
management strategies (5). In cases when coping fails, stress ensues,
often accompanied by physiological reactions like increased HR (6). As
part of the coping process, and dependent on its outcomes, teachers
re-evaluate the stressor (7).

As shown in Fig.\ref{fig.1}, teachers' primary and secondary appraisal
as well as coping attempts are influenced by professional experience. As
professional experience grows, teachers develop cognitive scripts for
managing classroom events, resulting in more complex and problem-focused
classroom management skills (Wolff, Jarodzka, and Boshuizen 2021), and
thus more effective coping and less stress. Empirically, classroom
management skills and problem-focused coping styles are linked to fewer
instances of emotional exhaustion (Maslach, Schaufeli, and Leiter 2001;
Clunies-Ross, Little, and Kienhuis 2008). Novices in the teaching
profession, on the other hand, face considerable stress and often feel
overwhelmed by the demands of teaching (Ophardt and Thiel 2017; Wolff et
al. 2015; Klusmann et al. 2012) with many leaving the profession within
the first five years (Ingersoll and Smith 2003). Accordingly, when
resources are lacking and coping fails, negative consequences for health
(e.g., burnout) and for work (e.g., high turnover rates) can arise
(Jalongo and Heider 2006; Unterbrink et al. 2007; Aloe et al. 2014),
highlighting the importance of professional expertise for managing
teacher stress (Fisher 2011).

@@

To measure teachers' HR, we used the wrist-based fitness tracker Fitbit®
Charge 4. In line with the manufacturer's instructions (Fitbit 2020),
the device was attached to the participants' nondominant hand, a
finger's width above the wrist bone. The tracker works by flashing green
LEDs hundreds of times per second, using light-sensitive photodiodes to
catch the reflected light, to calculate the volume changes in the
capillaries. From this, the tracker calculated the heart beats per
minute. HR measurements were generated at least every 15
seconds\footnote{The fluctuations in the number of seconds in which the
  HR was measured are due to the participants' movements, meaning that
  the device could not measure the HR every second.}. The raw data
contained the estimated HR in BPM for each time stamp. To account for
individual differences in the baseline HR, we also calculated
z-standardized HR values based on individual means, i.e., at the subject
level of \(n\) = 81 participants (standardized HR).

Since we aimed to keep measurement intervals comparable between study
phases, we aggregated HR over a representative 10-minute interval within
each phase (cf.~Fig.\ref{fig.2}). Previous research has indicated that
10-minute intervals are a useful duration for analyzing PPG data (Lu et
al. 2008). The intervals were selected based on the following rules: The
\emph{pre-teaching interval} (\(I_1\)) comprised the first 10 minutes
after the fitness tracker had been put on. The \emph{teaching interval}
(\(I_2\)) started two minutes after the lesson had started. This
interval was of the highest relevance to our study. We explicitly chose
an early 10-minute interval within the teaching phase, as previous
studies revealed that the beginning of a lesson is most demanding and
potentially stressful with regards to teacher-student interaction
(Donker, Van Gog, and Mainhard 2018; Claessens et al. 2017). The
\emph{post-teaching interval} (\(I_3\)) started immediately after the
end of the teaching unit. The \emph{interview interval} (\(I_4\)) was
defined as the mid-10 minutes between the end of the teaching unit and
the time point when the fitness tracker was taken off. All participants
were being interviewed during this interval. The end interval (\(I_5\))
comprised the last 10 minutes before the fitness tracker was taken off.

@@

We conducted all analyses with R (RStudio Team 2020). Graphics were
created using ggplot2 (Wickham 2016).

To enable visual inspection of HR trends, we displayed smoothed teacher
HR over the course of the recording.\footnote{The curve was smoothed
  using the geom\_smooth() function from the ggplot2 package in R
  (Wickham 2016) based on the smoothing method LOESS (Locally Estimated
  Scatterplot Smoothing). This method fits a polynomial surface
  determined by one or more numerical predictors, using local fitting.}
We visually compared unstandardized and standardized HR trends over the
two-hour recording period.\footnote{Note that the study exceeded the
  planned duration of two hours for a few participants. To avoid
  distortions when mapping the HR over the course of the study (see
  Fig.\ref{fig.3}), the endpoint was set at two hours for all
  participants, even though data from later time points was used in the
  \emph{end interval} for a few participants.} For all further analyses,
we used standardized rather than unstandardized HR values.

We averaged each person's standardized HR over each of the five selected
intervals\footnote{We used the mean standardized HR instead of the mean
  intercept as we wanted to explain the mean HR of the entire intervals
  and not the HR at the very beginning of the interval (\(x\) = 0).},
resulting in one measure per person per interval. To test Hypothesis 1a,
we initially conducted a one-way ANOVA with repeated measures as an
omnibus test and then tested the mean differences between the
\emph{teaching interval} (\(I_2\)) and the other four intervals by
planned contrasts and inspection of effect size \(d\) (Cohen 1988).

For testing Hypothesis 1b, concerning HR changes within each interval,
we first conducted a linear estimation of the increase or decrease in
standardized HR values over time for each participant. To this end, we
used fixed intercept fixed slope regression models (Gelman and Hill
2006) for each interval to estimate intercepts and linear slopes for
each individual, which were then averaged across individuals.\footnote{Although
  this procedure does not account for nonmonotonic progressions in
  individual HR, a graphical evaluation revealed that the linear
  estimates corresponded well to the majority of the cases (see
  Fig.\ref{fig.a3} to \ref{fig.a7} in the supplementary material).} We
tested Hypothesis 1b based on the unstandardized estimates of mean
slopes (one estimate per participant per interval).

@@

These findings are in line with prior studies showing that teachers' HR
varies depending on their activities and encountered stressors with
increases during phases where teachers are in an exposed position
(Sperka and Kittler 1995; Scheuch and Knothe 1997; Donker, Van Gog, and
Mainhard 2018; Junker, Donker, and Mainhard 2021), as well as with
findings showing how HR changes align with activating events and
stress-inducing tasks (Darnell and Krieg 2019; Chalmers et al. 2021).

Building on the model of teacher stress (Kyriacou and Sutcliffe (1978),
see Fig.\ref{fig.2}), we had hypothesized that more experienced
teachers, with better classroom management skills at their disposal,
experience less physiological stress when dealing with classroom
disruptions. Contrary to our expectations, we found no buffering effect
of teaching experience on teachers' HR, i.e., more experienced teachers
did not show lower mean standardized HR during the stressful teaching
phase than less experienced teachers. Rather, at least descriptively, we
observed the opposite trend. There are several possible explanations for
this finding. First, teaching experience is inherently confounded with
age (the two variables correlated at \(r = .94\) in our sample), and age
has been shown to affect indicators of cardiovascular reactivity in
various ways (Uchino, Birmingham, and Berg 2010). However, to avoid this
kind of confounding influence, we had used not raw BPM but rather
standardized mean HR for all our analyses, thus controlling at least for
inter-individual differences in mean HR. Second, as research on teacher
professionalization has repeatedly shown, professional experience is not
a guarantee for higher professional knowledge and skills (Kirschner et
al. 2016). Rather, honing skills from professional experience
necessitates a deliberate practice of choosing to improve, learning
through experience, and integrating new knowledge into future
performances (Dunn and Shriner 1999). Thus, rather than professional
experience alone, more direct assessments of classroom management
skills, such as objective behavior-based tests, would be a better
indicator of expertise that future studies could explore. Finally, and
most importantly, the highly controlled teaching situation that we
created in the lab might not have provided sufficient resemblance to the
expert teachers' working conditions to let them effectively use their
coping resources. In other words, since the situation was unfamiliar to
both experienced and unexperienced teachers, their stress levels might
have been more similar than they would have been in a more authentic
classroom setting.
